Leadership across cultures and contexts: an exploratory analysis

### 2020-04-29 leadership_across_cultures_contexts.Rmd
labeled intro paragraphs

Across human societies, the influence and agendas of leaders stands to substantially shape group structure and group outcomes. The structure of groups, their goals, and the demands of leaders can favor particular leadership qualities as well as the costs and benefits to individuals involved. Cultural norms, capacities of competing leaders, inter-group dynamics, and follower preferences can also impact leader agendas. The interactions between leader strategies, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories [@wiessner_risk_1982] and cultural and biological evolutionary processes [@henrich_big_2015].

@@

Much scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover the interrelationships of these dimensions of leadership. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020].

There is now well established consensus within behavioral and evolutionary human sciences on the importance of data from the full range of human cultural diversity and systemic biases in much of social science are well known [@henrich_weirdest_2010, *The weirdest people in the world?*]. @henrich_weirdest_2010 in their initial critique of the over-realiance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies, but also identified a contrast that, "compares people from modern industrialized societies with those from small-scale societies" (p.62), which has become the most salient interpretation of the paper and call-to-action. Since the seminal WEIRD people paper many social scientists have indeed taken up the challenge to diversify their samples and have sought out "non-WEIRD" participants and databases. However, categorizing populations as WEIRD or non-WEIRD and attempting to charachterize "non-WEIRD" social dynamics to learn about human psychology is a reductionist over-correction and arguably only a marginal scientific improvement from exclusively studying WEIRD populations.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] Extending the study of leadership to include evolutionary theory and data from diverse, small-scale and traditional societies (i.e., non-WEIRD) can facilitate more generalizable theories and offer applications to shape organizational structure across populations (i.e., WEIRD settings) in ways which are more consistent and less mismatched with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. But such approaches are also at risk of failing to incorporate the message of the larger WEIRD people problem: diversity of cultural, social, and ecological contexts can be strong divers of human behavior and group dynamics.

Much of leadership studies, including evolutionary approaches, have focused on community or political leaders, often overlooking subtilties in the nature of leadership and followership across social domains and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. Lastly, while the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated. Outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is associated with population-level variation and in different social contexts and identifying the associated benefits *and* costs both leaders and followers incur.
<!-- CAN PULL SOME COST CITATIONS FROM NHB PAPER -->

<!-- THIS PARAGRAPH CAN BE INCORPORATED ABOVE -->
Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. Leadership roles are also often associated with social sanctioned rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @von_rueden_leadership_2015].

We first briefly review theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study before presenting results and interpretations. Through novel exploratory analyses encompassing the full range of cultural diversity and diverse social contexts we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches to human leadership across cultures and contexts.

>>>>

<!-- INTRO TO MOTIVATE PAPER -->
Across human societies, the influence and agendas of leaders stands to substantially shape group structure and group outcomes. The structure of groups, their goals, and the demands of leaders can favor particular leadership qualities as well as the costs and benefits to individuals involved. Cultural norms, capacities of competing leaders, inter-group dynamics, and follower preferences can also impact leader agendas. The interactions between leader strategies, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories [@wiessner_risk_1982] and cultural and biological evolutionary processes [@henrich_big_2015].

@@

<!-- EVOLUTIONARY THEORY APPROACH -->
Much scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover the interrelationships of these dimensions of leadership. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020].

<!-- WEIRD PEOPLE PROBLEMS -->
There is now well established consensus within behavioral and evolutionary human sciences on the importance of data from the full range of human cultural diversity and systemic biases in much of social science are well known [@henrich_weirdest_2010, *The weirdest people in the world?*]. @henrich_weirdest_2010 in their initial critique of the over-realiance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies, but also identified a contrast that, "compares people from modern industrialized societies with those from small-scale societies" (p.62), which has become the most salient interpretation of the paper and call-to-action. Since the seminal WEIRD people paper many social scientists have indeed taken up the challenge to diversify their samples and have sought out "non-WEIRD" participants and databases. However, categorizing populations as WEIRD or non-WEIRD and attempting to charachterize "non-WEIRD" social dynamics to learn about human psychology is a reductionist over-correction and arguably only a marginal scientific improvement from exclusively studying WEIRD populations.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] Extending the study of leadership to include evolutionary theory and data from diverse, small-scale and traditional societies (i.e., non-WEIRD) can facilitate more generalizable theories and offer applications to shape organizational structure across populations (i.e., WEIRD settings) in ways which are more consistent and less mismatched with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. But such approaches are also at risk of failing to incorporate the message of the larger WEIRD people problem: diversity of cultural, social, and ecological contexts can be strong divers of human behavior and group dynamics.

<!-- GROUP CONTEXT and cost PROBLEMS -->
Much of leadership studies, including evolutionary approaches, have focused on community or political leaders, often overlooking subtilties in the nature of leadership and followership across social domains and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. Lastly, while the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated. Outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is associated with population-level variation and in different social contexts and identifying the associated benefits *and* costs both leaders and followers incur.
<!-- CAN PULL SOME COST CITATIONS FROM NHB PAPER -->

<!-- THIS PARAGRAPH CAN BE INCORPORATED ABOVE ????? -->
Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. Leadership roles are also often associated with social sanctioned rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @von_rueden_leadership_2015].

<!-- CLOSING -->
We first briefly review theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study before presenting results and interpretations. Through novel exploratory analyses encompassing the full range of cultural diversity and diverse social contexts we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches to human leadership across cultures and contexts.
